Verification of Stencil Computations with Spatial Specifications

### 2016-07-07 paper.tex
shorten caption
andy edits

\title{Verification of Stencil Computations with Spatial
  Specifications}
\authorinfo{Blind}{}{}
%\authorinfo{Dominic Orchard \and Mistral Contrastin
%\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}

@@

In this example, the access pattern is simple, static, and easily
understood. More complex stencil computations
are much more prone to error from simple lexical mistakes and
harder to understand. For example,
Figure~\ref{ref:navier-stokes-fragment} shows three lines from a
Navier-Stokes fluid simulator in which two arrays are read from with
different data access patterns, across two dimensions. The interaction
is much harder to understand, with the potential for the developer to
accidentally introduce an error via simple textual mistakes, for
example writing $\texttt{(i-1,j)}$ instead of $\texttt{(i+1,j)}$.

In practice, most stencil computations have a regular shape that can
be described simply and abstractly with a small set of coarse-grained
spatial descriptions. We introduce a simple specification language of
such descriptions, abstracting over the fine-grained detail of stencil
access patterns. For the Laplace example, our
inference procedure provides the following specification to line 3:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This describes that \texttt{a} is accessed with a
symmetrical pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Fig.~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Fig.~\ref{ref:navier-stokes-fragment}(c) illustrates
the specification pictorially.

@@

\item we detail inference, checking, and specification synthesis 
algorithms for our language (Sect.~\ref{sec:analysis}), derived from
the a denotational model for the language (Sect.~\ref{sec:semantics});

@@

\item we perform a quantitative study of
  stencil computations in a corpus of numerical Fortran programs
  (ranging from small to large),  totalling one million lines of code. Our tool identifies and
  synthesises specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors;

@@

Our specification language is based on the observation
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
``\emph{five-point stencil}'' on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in Section~\ref{sec:evaluation} with the inference of
such regular stencil patterns on our corpus of numerical programs.

@@

\renewcommand*{\arraystretch}{0.8}
For the target language, $e$ ranges over expressions (may be impure) and $v$
overs its (imperative) variables. %By
%``variable'', we mean in the imperative sense (named binders to memory cells).

\begin{definition}[Base induction variable]
  A variable \textit{v} of integer type is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop,\footnote{Or equivalent, \eg{}, \texttt{do} in Fortran, for our implementation.} incremented by $1$ per
iteration. The variable is marked as an induction variable
only within the body of the loop. Since we ignore non-base
(derived) induction variables, we simply say \emph{induction
  variable}. Induction variables are ranged over by $i$, $j$, $k$ throughout. 
\end{definition}

\begin{definition}[Array subscript and index]
  An \emph{array subscript} is an expression that indexes an array
  (for the purpose of reading or writing an element), which we denote
  by $v(\bar{e})$ for the source language where $\bar{e}$ is shorthand
  for a syntactic list of indexing expressions. A \emph{relative
    index} is a list $\bar{e}$ where each $e \in \bar{e}$ is defined
  in terms of base induction variables.
\end{definition}

\begin{definition}[Neighbourhood index]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is an
  integer constant. That is, a neighbourhood index is a
  relative index defined as constant translations of base induction
  variables. (The relation $\equiv$ here identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ \eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices 
  by the predicate $\neigh_I$ where $I$ is a set of induction
  variables. We omit $I$ when implicit from the context.
\label{def:neighbour}
\end{definition}

@@

\emph{Regions} are the central building blocks of spatial
specifications. Regions can either be declared along with a
\nonterm{regionDec}, assigning a region specification \nonterm{region} to
a region variable \nonterm{rvar} or given directly within a
\nonterm{spatial} specification.

@@

Regions have as terminals the \emph{region constants}, denoted by
\term{reflexive}, \term{forward}, \term{backward}, or \term{centered}. Each
region constant except \term{reflexive} is given a depth parameter $n$ (natural
number greater than 0). Every region constant receives a dimension identifier $d$ (also a
natural number greater than 0). Region constants specify that an array is read
from by a collection of indices which in the $d^{th}$ dimension are
neighbourhood indices ranging from $i + 0$ up to $i \pm n$ inclusively,
\eg{}, the following is a valid stencil, reading from \term{a} and writing to
\term{b}:
%%
\begin{minted}{fortran}
!= stencil forward(depth=2, dim=1) :: a
b(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
\end{minted}
%%
were \texttt{i} is an induction variable.  The specification
is associated to the array variable \texttt{a}. Note that the second
dimension is constant.

@@

The product of two regions $r \texttt{*} r'$ specifies that an array
is read with neighbourhood indices drawn from the bounding box created
by the regions two regions $r$ and $r'$. For example, the following
code (defining a \emph{nine-point stencil}) has a specification given by the
product of two \texttt{centered} regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%%
This pattern is a common image convolution (for example
for edge detection). The specification ranges over the
values that flow the array subscript on the left-hand side,
and so ranges over the intermediate assignments to \term{x},
\term{y}, and \term{z}.

The region constructor \texttt{*} can also be interpreted as a
kind of conjunction on specifications since they constraint
indices to be within both regions simultaneously.

@@

\begin{figure}
%\vspace{-0.2em}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.6em]
\begin{array}{rll}
(\textsc{F\, \texttt{+} \,F}) \;\; &
\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,B}) \;\; &
\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,C}) \;\; &
\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,B}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,F}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
  \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,F}) \;\; &
\stenFwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,B}) \;\; &
\stenBwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,C}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                         (R^\irreflS \, \term{*} \, S) \;
\equiv \; R \, \term{*} \, S \\[0.4em]
%
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \quad
%
(\texttt{+}\textsc{comm}) \;\; S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.4em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.4em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)
\end{array}
%
%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
%{\textit{specDec'} :: \texttt{v}}$} \\[0.7em]
%\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
%\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications (\emph{regions})}
\vspace{-1em}
\label{fig:equations}
\end{figure}

@@

\begin{figure}[t]
\begin{align*}
\hspace{-0.5em}
\begin{array}{c}
%%
%% INEQUATIONS ON REGION
%% 
%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
%
%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
%  Q}\trule{\textsc{trans}} 
%\qquad
%\\[1.5em]
\vspace{-0.4em}
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} m \leq n \qquad\qquad r \Rightarrow s \hspace{3em}}
{\begin{array}{rl}
\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
\end{array}}\trule{\textsc{over}} \\[-0.2em]
\hspace{-0.5em}
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{+} \, P_2 <:^r R_1 \, \texttt{+} \, R_2}
\trule{\textsc{cong}\texttt{+}} \;\;\;
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{*} \, P_2 <:^r R_1 \, \texttt{*} \, R_2}
\trule{\textsc{cong}\texttt{*}} \\[1.5em]
%%
%% INEQUATIONS ON SPEC
%%
%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
%
\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
\quad
\dfrac{}{\term{atLeast} \, R <: \term{atMost} \,
  R}\trule{\textsc{bound}} \\[1.5em]
%%
\dfrac{P <:^r R}{\term{atLeast}  P <: \term{atLeast} 
  R}\trule{\textsc{cong}\texttt{L}}
\ 
\dfrac{P <:^r R}{\term{atMost}  P <: \term{atMost} 
  R}\trule{\textsc{cong}\texttt{M}} \\[1.5em]
\dfrac{P <:^r R}{\term{readOnce} \, P <: \term{readOnce} \,
  R}\trule{\textsc{cong}\texttt{R}} 
\quad
\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}} 
%%
\end{array}
\end{align*}
\caption{Inequations $<:^r$ on \textit{regions} and
$<:$ on \textit{specs} (omitting reflexivity
and transitivity). }
\label{fig:inequations}
\vspace{-0.8em}
\end{figure}

@@

\begin{example}
The following three array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$ 
are given by \textsf{schematic} the 
index schemes $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
and $\vtwoh{\infty}{0}$ respectively. The following
relationship of generalises then holds:
%\[
%\xymatrix@R=0.2em@C=2em{
%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
%& [0 \;\; 1]  \\
%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} &
%}
%\]
%%
$\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1}$
and
$\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1}$.

As an example of undefined \textsf{schematic}, 
 $\textsf{schematic}(i*2, j) = \bot$ as its first index is a non-neighbour relative index, which
 is outside the scope of our specification language.
\end{example}
\noindent
Our terminology of \emph{indexing schemes} and
\emph{generalisation} is by analogy
with the \emph{type schemes} and \emph{type generalisation} of
polymorphic types in ML~\cite{milner1978theory}. Our notion is
much more rigid however, as our schemes do not bind names
and thus generalisation does not require substitution; $\infty$ acts
as a positional placeholder.

From \textsf{schematic} we now give a
formal characterisation of stencil computations for our purposes:
%%
\begin{definition}[Stencil computations]
  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
  collection of arrays of arbitrary (possible differing)
  dimensionalities.  
  A \emph{stencil computation} comprises an iteration by a set of at
  most $n$ induction variables $I$ 
  over a subset of the index space of $a$. Elements of $a$ are determined by an
  assignment $a(\bar{e}) = e_r$ where $\bar{e}$ is an array index such
  that  $\mathsf{schematic}_I(\bar{e}) \neq \bot$ and for all array
  subscripts $b(\bar{e}') \in e_r$ then
  $\mathsf{schematic}_I(\bar{e}') \neq \bot$. 
\label{def:stencil}
\end{definition}
%%

@@

\subsubsection{Step 2: Data-access analysis}
\label{sec:inf-step2}

@@

>>>>

\title{Verification of Stencil Computations with Spatial
  Specifications}
\authorinfo{}{}{}
%\authorinfo{Dominic Orchard \and Mistral Contrastin
%\and Matthew Danish \and Andrew Rice}{University of Cambridge}{}

@@

In this example, the access pattern is simple, static, and easily
understood. More complex stencil computations
are much more prone to error from simple lexical mistakes. For example,
Figure~\ref{ref:navier-stokes-fragment} shows three lines from a
Navier-Stokes fluid simulator in which two arrays are read with
different data access patterns, across two dimensions. The interaction
is much harder to understand, with the potential for the developer to
introduce an error via simple textual mistakes, for
example writing $\texttt{(i-1,j)}$ instead of $\texttt{(i+1,j)}$.

In practice, most stencil computations have a regular shape that can
be described simply and abstractly with a small set of coarse-grained
spatial descriptions. We introduce a simple specification language of
such descriptions, abstracting over the fine-grained detail of stencil
access patterns. For the Laplace example, our
inference procedure provides the following specification to line 3:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
This describes that \texttt{a} is accessed with a
symmetrical pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
The inferred specification for the Navier-Stokes example is shown 
in Fig.~\ref{ref:navier-stokes-fragment}(b). The
specification requires that, over the whole fragment, \texttt{u} is
accessed with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five-point stencil}) and \texttt{v}
is accessed in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Fig.~\ref{ref:navier-stokes-fragment}(c) illustrates
the specification pictorially.

@@

\item we detail inference, checking, and specification synthesis 
algorithms for our language (Section~\ref{sec:analysis}), derived from
a denotational model for the language (Section~\ref{sec:semantics});

@@

\item we report on a quantitative study of
  stencil computations on a corpus of numerical Fortran programs
  (ranging from small to large),  totalling one million lines of code. Our tool identifies and
  synthesises specifications for 35,000 stencil computations in the
  corpus. Approximately 10\% of the stencils we found are non-trivial,
  corresponding to code which is a possible source of errors;

@@

Our specification language is based on the hypothesis
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
\emph{five-point stencil} on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in Section~\ref{sec:evaluation} with the inference of
such regular stencil patterns on our corpus of numerical programs.

@@

\renewcommand*{\arraystretch}{0.8}
For the target language, $e$ ranges over expressions (may be impure) and $v$
over its (imperative) variables. %By
%``variable'', we mean in the imperative sense (named binders to memory cells).

\begin{definition}[Base induction variable]
  A integer variable \textit{v} is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop,\footnote{Or equivalent, \eg{}, \texttt{do} in Fortran, for our implementation.} incremented by $1$ per
iteration. The variable is marked as an induction variable
only within the body of the loop. Since we ignore non-base
(derived) induction variables, we simply say \emph{induction
  variable}. Induction variables are ranged over by $i$, $j$, $k$ throughout. 
\end{definition}

\begin{definition}[Array subscript and index]
  An \emph{array subscript} is an expression that indexes an array, which we denote
  by $v(\bar{e})$ for the source language where $\bar{e}$ is shorthand
  for a syntactic list of indexing expressions. A \emph{relative
    index} is a list $\bar{e}$ where each $e \in \bar{e}$ is defined
  in terms of base induction variables.
\end{definition}

\begin{definition}[Neighbourhood index]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is an
  integer constant. That is, a neighbourhood index is a
  relative index defined as constant translations of base induction
  variables. (The relation $\equiv$ here identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ \eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices 
  by the predicate $\neigh_I$ where $I$ is a set of induction
  variables. We omit $I$ when clear from the context.
\label{def:neighbour}
\end{definition}

@@

\emph{Regions} are the central building blocks of spatial
specifications. Regions can either be declared alone with a
\nonterm{regionDec}, assigning a region specification \nonterm{region} to
a region variable \nonterm{rvar} or given directly within a
\nonterm{spatial} specification.

@@

Regions have as terminals the \emph{region constants}, denoted by
\term{reflexive}, \term{forward}, \term{backward}, or \term{centered}. Each
region constant except \term{reflexive} is given a depth parameter $n$ (natural
number greater than 0). Every region constant receives a dimension identifier $d$ (also a
natural number greater than 0). Region constants specify that an array is read
from by a collection of indices which in the $d^{th}$ dimension are
neighbourhood indices ranging from $i + 0$ up to $i \pm n$ inclusively,
\eg{}, the following is a valid stencil, reading from \term{a} and writing to
\term{b}:
%%
\begin{minted}{fortran}
!= stencil forward(depth=2, dim=1) :: a
b(i, 0) = a(i, 0) + a(i+1, 0) + a(i+2, 0)
\end{minted}
%%
where \texttt{i} is an induction variable.  The specification
is associated to the array variable \texttt{a}. Note that the second
dimension is constant.

@@

The product of two regions $r \texttt{*} r'$ specifies that an array
is read with neighbourhood indices drawn from the bounding box created
by the regions two regions $r$ and $r'$. For example, the following
code (defining a \emph{nine-point stencil}) has a specification given by the
product of two \texttt{centered} regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%%
This pattern is common in image convolution applicaitons (for example
for edge detection). The specification ranges over the
values that flow to the array subscript on the left-hand side,
and so ranges over the intermediate assignments to \term{x},
\term{y}, and \term{z}.

The region constructor \texttt{*} can also be interpreted as a
kind of conjunction on specifications since they constrain
indices to be within both regions simultaneously.

@@

\begin{figure}
%\vspace{-0.2em}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.6em]
\begin{array}{rll}
(\textsc{F\, \texttt{+} \,F}) \;\; &
\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,B}) \;\; &
\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,C}) \;\; &
\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,B}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,F}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
  \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,F}) \;\; &
\stenFwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,B}) \;\; &
\stenBwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,C}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                         (R^\irreflS \, \term{*} \, S) \;
\equiv \; R \, \term{*} \, S \\[0.4em]
%
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.4em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.4em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)
\end{array}
%
%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
%{\textit{specDec'} :: \texttt{v}}$} \\[0.7em]
%\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
%\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications (\emph{regions})}
\vspace{-0.5em}
\label{fig:equations}
\end{figure}

@@

\begin{figure}[t]
\begin{align*}
\hspace{-0.5em}
\begin{array}{c}
%%
%% INEQUATIONS ON REGION
%% 
%\framebox{$\textit{region} <: \textit{region}'$} \\[0.85em]
%
%\dfrac{}{R <:^r R}\trule{\textsc{refl}} \qquad \dfrac{P <:^r R \quad R <:^r Q}{P <:^r
%  Q}\trule{\textsc{trans}} 
%\qquad
%\\[1.5em]
\vspace{-0.4em}
\setlength{\arraycolsep}{0.1em}
\dfrac{\hspace{3em} m \leq n \qquad\qquad r \implies s \hspace{3em}}
{\begin{array}{rl}
\stenFwdSR{m}{d}{r} & <:^r \stenFwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenBwdSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenCenSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s}
%& \;\, ds \subseteq es \wedge n \leq m
\\
\wedge \; \stenBwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\
\wedge \; \stenFwdSR{m}{d}{r} & <:^r \stenCenSR{n}{d}{s} \\[1.2em]
\end{array}}\trule{\textsc{over}} \\[-0.2em]
\hspace{-0.5em}
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{+} \, P_2 <:^r R_1 \, \texttt{+} \, R_2}
\trule{\textsc{cong}\texttt{+}} \;\;\;
\dfrac{P_1 <:^r R_1 \quad P_2 <:^r R_2}
      {P_1 \, \texttt{*} \, P_2 <:^r R_1 \, \texttt{*} \, R_2}
\trule{\textsc{cong}\texttt{*}} \\[1.5em]
%%
%% INEQUATIONS ON SPEC
%%
%\framebox{$\textit{spec} <: \textit{spec}'$} \\[0.8em]
%
\dfrac{}{\term{readOnce} \, R <: R}\trule{\textsc{rep}}
\quad
\dfrac{}{\term{atLeast} \, R <: \term{atMost} \,
  R}\trule{\textsc{bound}} \\[1.5em]
%%
\dfrac{P <:^r R}{\term{atLeast}  P <: \term{atLeast} 
  R}\trule{\textsc{cong}\texttt{L}}
\ 
\dfrac{P <:^r R}{\term{atMost}  P <: \term{atMost} 
  R}\trule{\textsc{cong}\texttt{M}} \\[1.5em]
\dfrac{P <:^r R}{\term{readOnce} \, P <: \term{readOnce} \,
  R}\trule{\textsc{cong}\texttt{R}} 
\quad
\dfrac{R \equiv S}{R <:^r S}\trule{\textsc{eq}} 
%%
\end{array}
\end{align*}
\caption{Inequations $<:^r$ on \textit{regions} and
$<:$ on \textit{specs} (omitting reflexivity
and transitivity). }
\label{fig:inequations}
\vspace{-0.8em}
\end{figure}

@@

\begin{example}
The following three array indexing terms
$(i, j+1)$, $(1, j+1)$ and $(3, j)$ 
are mapped by \textsf{schematic} to the 
index schemes $\vtwoh{0}{1}$, $\vtwoh{\infty}{1}$
and $\vtwoh{\infty}{0}$ respectively. The following
relationships of generalisation then holds:
%\[
%\xymatrix@R=0.2em@C=2em{
%[\infty \;\; 0] \ar@{-}[dr]^{\preceq_n} &  \\
%& [0 \;\; 1]  \\
%[\infty \;\; 1]  \ar@{-}[ur]_{\preceq_n} &
%}
%\]
%%
$\vtwoh{\infty}{0} \; \not\preceq_2 \; \vtwoh{0}{1}$
and
$\vtwoh{\infty}{1} \; \preceq_2 \; \vtwoh{0}{1}$.

As an example of undefined \textsf{schematic}, 
 $\textsf{schematic}(i*2, j) = \bot$ as its first index is a non-neighbour relative index, which
 is outside the scope of our specification language.
\end{example}
\noindent
Our terminology of \emph{indexing schemes} and
\emph{generalisation} is by analogy
with the \emph{type schemes} and \emph{type generalisation} of
polymorphic types in ML~\cite{milner1978theory}. Our notion is
much more rigid however, as our schemes do not bind names.
%and thus generalisation does not require substitution; $\infty$ acts
%as a positional placeholder.

From \textsf{schematic} we now give a
formal characterisation of stencil computations for our purposes:
%%
\begin{definition}[Stencil computations]
  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a
  collection of arrays of arbitrary (possible differing)
  dimensionalities.  
  A \emph{stencil computation} comprises an iteration by a set $I$ of at
  most $n$ induction variables 
  over a subset of the index space of $a$. Elements of $a$ are determined by an
  assignment $a(\bar{e}) = e_r$ where $\bar{e}$ is an array index such
  that  $\mathsf{schematic}_I(\bar{e}) \neq \bot$ and for all $b \in
  \bar{b}$ and all subscripts $b(\bar{e}') \in e_r . \, $
  $\mathsf{schematic}_I(\bar{e}') \neq \bot$. 
\label{def:stencil}
\end{definition}
%%

@@

\subsubsection{Step 2: Analyse data-access}
\label{sec:inf-step2}

@@

\subsection{Approximations and checking}
